---
title: "Introduction to rspeaksnonmem"
author: "Mike K Smith (Mike.K.Smith@Pfizer.com)"
date: "`r Sys.Date()`"
#output: rmarkdown::html_vignette
output: html_notebook
---

```{r setup,echo=FALSE,results='hide',message=FALSE, warning=FALSE}
 devtools::load_all(pkg = ".")
file.copy(from = file.path(system.file("exdata", package = "rspeaksnonmem"),
                           "warfarin_conc_pca.csv"),
            to = getwd(), overwrite = T )
file.copy(from = file.path(system.file("exdata", package = "rspeaksnonmem"),
                           "warfarin.ctl"),
            to = getwd(), overwrite = T )
file.rename(from = "warfarin.ctl", "run0.ctl")
```

In this notebook we will look at running NONMEM and PsN from R using the R
package `rspeaksnonmem`. 

### Overview

This demo will show the following steps:
  
  - Running NONMEM and PsN
  - Integrating results into a Notebook.
  
### Step 1: Running NONMEM
First, let's run a model and knit the results back into this notebook. Note 
that in the `command` argument to the function we specify the complete path to the
NONMEM executable .bat file. This allows you to be very precise about which
version of NONMEM you use. If the `clean` argument is non-zero it cleans up the 
working directory deleting all of the temporary NONMEM files, leaving only the 
output. 

```{r RunNONMEM}
  estimate_NM(command = "c:/nm72/run/nmfe72.bat",
              modelFile = "run0.ctl",
              clean = 1)
```

### Step 2: Summarise estimation:
We can now use PsN command `sumo` to summarise the output. Note that here we
are not using the `command` argument, but instead we specify the installation
directory for Perl / PsN and which PsN version we are using. `rspeaksnonmem` 
will then create a command based on the directory, `sumo` and version information.

```{r callSumo}
  sumo_PsN(installPath = "c:/strawberry/perl",
           version = "4.7.0",
           lstFile = "run0.lst")
```
### Step 3: Model Diagnostics
Next, we might want to produce some model diagnostics using the R package 
`xpose4`. To do this, we need to create an Xpose database object and then 
create the diagnostic plots. Xpose needs the run number in order to pick up
the $TABLE file outputs from NONMEM.

```{r setupXposeDiagnostics, warning=FALSE, message=FALSE}
  runno <- as.numeric(gsub("[a-z]", "", list.files(pattern = "^sdtab")[1]))
  
  ## ----createXpdb----------------------------------------------------------
  xpdb <- xpose4::xpose.data(runno, quiet = T)
  # save(base.xpdb, file='Xpose database.RData')
```

#### DV vs PRED / IPRED
```{r dv.vs.pred.ipred, warning=FALSE, message=FALSE}
  print(xpose4::dv.vs.pred.ipred(xpdb))
```
#### PRED vs IDV
```{r pred.vs.idv, warning=FALSE, message=FALSE}
  print(xpose4::pred.vs.idv(xpdb))
```
  
#### PRED vs IDV
```{r ipred.vs.idv, warning=FALSE, message=FALSE}
  print(xpose4::ipred.vs.idv(xpdb))
```
#### WRES vs IDV
```{r wres.vs.idv, warning=FALSE, message=FALSE}
  print(xpose4::wres.vs.idv(xpdb))
```
#### WRES vs PRED
```{r wres.vs.pred, warning=FALSE, message=FALSE}
  print(xpose4::wres.vs.pred(xpdb))
```

#### EBEs histogram
```{r ranpar.hist, warning=FALSE, message=FALSE}
  print(xpose4::ranpar.hist(xpdb))
```

#### Individual Plots
```{r ind.plots, warning=FALSE, message=FALSE}
  print(xpose4::ind.plots(xpdb, layout = c(4, 4)))  
```

### Step 4: VPC diagnostisc
We may also want to investigate simulation-based diagnostics using the PsN
routine VPC. In the call to `VPC_PsN` we have specified the required arguments
for VPC (samples, seed) but you can also pass additional arguments to the VPC
routine by passing a name-value list to the argument `psnOpts`. `rspeaksnonmem`
will check these arguments against the options for VPC and let you know if any
are mis-specified. Here the argument `min_points_per_bin` is not a valid option
for VPC. (The correct option is `min_points_**in**_bin`.)

```{r VPC_PsN}
VPC_PsN( modelFile='run0.ctl', 
         installPath = "c:/strawberry/perl",
         version = "4.7.0",
         working.dir='Run1_VPC', 
         samples=100, seed='54321',
         psnOpts=list(lst='run0.lst', 
                      auto_bin = '5',
                      min_points_per_bin = '2'))
```

and then use the Xpose function xpose_vpc to plot the result:
```{r plot_VPC}
vpcdir <- "Run1_VPC"

vpctab <- file.path(vpcdir,"vpctab")
vpcinfo <- file.path(vpcdir,"vpc_results.csv")

xpose4::xpose.VPC(vpc.info=vpcinfo,	vpctab=vpctab)
```