---
title: "rspeaksnonmem with PsN rplots .Rmd files"
author: "Mike K Smith"
date: "03 July 2017"
output: html_document
params: 
  rplots: 1
  psnRunDir: "run1"
---

```{r setup, message = FALSE}
# devtools::install_github(repo = "MikeKSmith/rspeaksnonmem")
# devtools::install_github(repo = "MangoTheCat/RNMImport")

library(rspeaksnonmem)
setwd(".")
```

```{r copyFiles,echo=FALSE,results='hide',message=FALSE, warning=FALSE}
file.copy(from = file.path(system.file("exdata", package = "rspeaksnonmem"),
                           "warfarin_conc_pca.csv"),
            to = getwd(), overwrite = T )
file.copy(from = file.path(system.file("exdata", package = "rspeaksnonmem"),
                           "warfarin.ctl"),
            to = getwd(), overwrite = T )
file.rename(from = "warfarin.ctl", "run0.ctl")
```

Being able to run PsN from within R means that you can easily combine execution
with the PsN output plot markdown files (.Rmd files within the PsN working 
directory). Here we're going to run the PsN `execute` function with the option
to create diagnostic plots `rplots=1`. 

We have parameterised this report by including the following parameters in the 
YAML header for this report:
```
---
title: "rspeaksnonmem with PsN rplots .Rmd files"
author: "Mike K Smith"
date: "03 July 2017"
output: html_document
params: 
  rplots: 1
  psnRunDir: "run0"
---
```

We can then use these parameters within the R code by referencing params$rplots
or params$psnRunDir. Having these as parameters of the document means that we
can re-run the document and change these parameters without having to change
code ***within*** the document.

## Run Execute
```{r PsN_execute_with_plots, message = FALSE, warning = FALSE}
execute_PsN(installPath = "c:/strawberry/perl",
            version = "4.7.0",
            modelFile = 'run0.ctl',
            rplots = params$rplots,
            working.dir = params$psnRunDir)
```

We search within the working directory for the .Rmd file and include this as
a child document to this parent markdown file. `rplotOutput` contains the name
of the PsN `rplots` output .Rmd file. In case we change the option above, we
also create a boolean variable `includeRplotOutput`.

```{r, results = "hide"}
 rplotOutput <- list.files(path = params$psnRunDir,
                           recursive = TRUE,
                           pattern = "plots.Rmd",
                           full.names = TRUE)
 includeRplotOutput <- params$rplots > 0 
```
We use the boolean `includeRplotOutput` to conditionally execute the chunk below.

```{r, child=rplotOutput, eval = includeRplotOutput}
```
