---
title: "Using rspeaksnonmem to modify NONMEM control streams"
author: "Mike K Smith (Mike.K.Smith@Pfizer.com)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`rspeaksnonmem` is designed to allow the user to craft workflows based on a 
given NONMEM model.

For more information on how to use `rspeaksnonmem` to run NONMEM and PsN please 
review the vignette "Using rspeaksnonmem to run NONMEM and PsN".

After importing and parsing a control stream to an R object using the 
`RNMImport` package, `rspeaksnonmem` allows the user to easily change initial 
estimates, change data attributes or change task properties (estimation or 
simulation settings) without having to change the model. This saves a lot of 
textual searching by the user. The resulting data, parameter,
model and task information is then written back out to file using an existing 
(possibly the original) model as a template. This ensures that the written 
NONMEM control stream mirrors as closely as possible the structure of the 
original file.

## Installation 

### Dependencies 

`rspeaksnonmem` relies on the package [`RNMImport`](https://github.com/MangoTheCat/RNMImport) 
function `importNmMod` which reads and parses the NONMEM control stream. 
`rspeaksnonmem` then works with the data, parameter values, and task information
separately from the model.

Install RNMIMport from Github using the `devtools` package. `rspeaksnonmem` 
requires the version of RNMImport on MikeKSmith's Github repository which 
includes handling of OMEGA and SIGMA blocks and identifying which elements are
fixed for these parameters.

```{r install_RNMImport, eval = FALSE}
  #devtools::install_git("MangoTheCat/RNMImport")
  #devtools::install_git("MikeKSmith/RNMImport")
```

### Install rspeaksnonmem
Eventually, `rspeaksnonmem` will be released to CRAN, but while still in 
development `rspeaksnonmem` can most easily be installed from GitHub using the 
`devtools` package:

```{r install_rspeaksnonmem, include=F}
    #devtools::install_github("MikeKSmith/rspeaksnonmem")
```


```{r,echo=FALSE,results='hide',message=FALSE}
 devtools::load_all(pkg = ".")
 devtools::load_all(pkg = "C:\\Users\\smith_mk\\Documents\\Working documents\\RNMImport")
```

Load the `rspeaksnonmem` package
```{r, include=FALSE, echo=FALSE}
#library(rspeaksnonmem)
```

Copy an example dataset and model to a directory of your choice

```{r,results='hide',echo=T}
getwd()

file.copy(from = file.path(system.file("exdata", package="rspeaksnonmem"),"warfarin_conc_pca.csv"),
            to = getwd(), overwrite=T )
file.copy(from = file.path(system.file("exdata", package="rspeaksnonmem"),"warfarin.ctl"),
            to = getwd(), overwrite=T )
```

The initial model within the workflow should act as a "template" for 
modifications. The best practice is to use a model where all possible parameters 
are defined (including OMEGAs and covariances / correlations) but where it is 
possible to fix parameters to zero or some null value. We can then run and test 
a wide variety of models simply by updating the $THETA, $OMEGA and $SIGMA 
parameters to allow estimation.

## Read the control stream using RNMImport 
First, we need to read the NONMEM control stream into R using the importNmMod 
function of RNMImport.
```{r readControlStream}
warfModel <- importNmMod("warfarin.ctl")
class(warfModel)
names(warfModel)
```

This creates a list of R objects containing the Raw NONMEM control stream as a 
vector of characters:
```{r viewRawNM}
warfBaseRaw <- warfModel$Raw
cat(warfModel$Raw, sep="\n")
```

It also creates a parsed list of control statements in the controlFile slot:
```{r viewParsedNM}
warfTemplateModel <- warfModel$problemContents[[1]]
warfModel$problemContents[[1]]
```

Using `rspeaksnonmem` the user can then update elements of the model using this 
parsed set of commands. `rspeaksnonmem` provides some additional functions to 
extract certain elements of the parsed control stream. 


"Object" | `rspeaksnonmem` Function  | NMTRAN blocks
---------|---------------------------|-----------------------------
Data     | getNMDataObjects          | `$DATA`, `$INPUT`
Parameter| getNMParameterObjects     | `$THETA`, `$OMEGA`, `$SIGMA`
Task     | getNMTaskPropertiesObjects| `$EST`,`$COV`, `$TAB`
Model    | getNMModelObjects         | everything else


For example: `getNMParameterObjects` returns the `$THETA`, `$OMEGA` and `$SIGMA` 
records. Since the `THETA` parameters are represented as a data frame, we can
easily update the initial values, lower or upper bounds, fix or estimate (unfix)
`THETA` parameters. With each object returned by the various `getNM<...>` 
functions we get not only the parsed objects but the raw control stream as well.

NOTE: Generally speaking, `rspeaksnonmem` is set up to allow the user to quickly
and easily update Data, Parameter or Task items, while leaving the Model 
unchanged. 

The `rspeaksnonmem` function `getNMDataObjects` returns the RAW `$DATA` and `$INPUT`
lines from the NONMEM control stream and the associated parsed objects.

```{r getNMData}
getNMDataObjects(warfModel)
```
 
Similarly, `getNMParameterObjects` returns the $THETA, $OMEGA and $SIGMA records.
```{r getNMParameterObjects}
getNMParameterObjects(warfModel)
```
 
Finally, `getNMTasProperties` returns the $EST, $COV and $TABLES records.
```{r getNMTaskPropertiesObjects}
getNMTaskPropertiesObjects(warfModel)
```

Together the Data, Parameters, Model and Task information forms a "Modelling 
Object Group" or MOG which is used for a specific estimation task.
We can update the MOG using the `rspeaksnonmem` function `updateModel`.

```{r updateMOG}
newTheta <- warfTemplateModel$Theta
newTheta["POP_TLAG",] <- list(Lower=0, Est=0.75, Upper=1.5, FIX=FALSE, comments="POP_TLAG")
newWarfModel <- updateModel(parsedObject = warfTemplateModel, theta=newTheta)
newWarfModel
```

Note how the THETA settings for POP_TLAG have been updated, but all other 
elements are unchanged. It is also possible to change the initial estimates 
while retaining the specified lower and upper bounds:

```{r}
newThetas <- warfTemplateModel$Theta
newInits <- c(0.25, 10, 4, 0.25, 0, 0.1)
newThetas[,"Est"] <- newInits

newModel <- updateModel(parsedObject = warfTemplateModel, theta=newThetas, runno=1)
newModel
```

We can finally write out the updated control stream using the function 
`writeNMControlStream`. This function uses an existing NONMEM control stream 
(or raw records) as a template for writing the updated, parsed statements. This 
ensures that the order of statements corresponds between the original and 
updated files, which allows easier checking of changes.

```{r writeNMControlStream}
writeNMControlStream(templateModel = warfModel$Raw, 
                     parsedControl = list(newModel), 
                     outputFile = "warfarin2", 
                     outFileExtension = "ctl")
cat(readLines("warfarin2.ctl", warn = F), sep="\n")
```

