---
title: "Using rspeaksnonmem"
author: "Mike Smith, Andrew C. Hooker and Ron Keizer "
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`rspeaksnonmem` is designed to allow the user to craft workflows based on a given NONMEM model.

`rspeaksnonmem` allows the user to run NONMEM models from R either directly running the nmfe.bat or by
calling [Perl speaks NONMEM](http://psn.sourceforge.net) (PsN) functions like "execute". Other PsN functions can be run from
R - VPC (Visual Predictive Check), bootstrap, SSE (Stochastic Simulation and Estimation).

After importing and parsing a control stream to an R object using the `RNMImport` package, `rspeaksnonmem` allows the user to easily change initial estimates, change data attributes or change task properties (estimation or simulation settings) without having to change the model. This saves a lot of textual searching by the user. The resulting data, parameter,
model and task information is then written back out to file using an existing (possibly the original) model as a 
template. This ensures that the written NONMEM control stream mirrors as closely as possible the structure of the original file.

## Installation

Eventually, `rspeaksnonmem` will be released to CRAN, but while still in development `rspeaksnonmem` can most easily be installed from GitHub using the `devtools` package:

    library(devtools)
    install_github("MikeKSmith/rspeaksnonmem")

## Dependencies

`rspeaksnonmem` relies on the package [`RNMImport`](https://r-forge.r-project.org/R/?group_id=1922) function 
`importNmMod` which reads and parses the NONMEM control stream. `rspeaksnonmem` then works with
the data, parameter values, and task information separately from the model.

Before installation of `rspeaksnonmem`, the package `RNMImport` needs to be installed. This package is not on CRAN but is available on R-Forge. You can install from there using:

    install.packages("RNMImport", repos="http://R-Forge.R-project.org")

or download the `.tar` file from [R-Forge](https://r-forge.r-project.org/R/?group_id=1922) and run in R:

    install.packages("~/Downloads/RNMImport_4.0-27.tar.gz", repos=NULL, type = "source")


## getting started
In order to execute NONMEM or PsN `rspeaksnonmem` needs to know where these tools are installed on your system.
The function defineExecutable creates the system command line for each tool by either referencing the tool name against 
an `installInfo` configuration file consisting of tool name, installation path and command e.g. "nonmem", "c:/nm72", "nmfe72.bat"; alternatively by searching for whether a command is configured on the system path; or by simply searching for executable files below an installation directory.

On Linux systems the "command" argument should provide the full path to the executable or define the command applicable
for running the tool at the system prompt. On Linux systems the "path" information on the configuration file may be left 
blank.

```{r setExecutables}
perlPath <- "c:/perl516"
PsNVersion <- "3.5.4"
NONMEMcommand <- "nmfe72"

tools <- c("NONMEM", "execute", "VPC", "bootstrap", "SSE", "runrecord", "sumo")
installPaths <- c("c:/nm72", rep(perlPath,length(tools)-1))
commands <- sapply(tools, function(x) ifelse( x == "NONMEM", NONMEMcommand, paste( x, PsNVersion, sep = "-" ) ) )

installedSoftware <- data.frame( cbind( tool = tools, path = installPaths, command = commands ), 
                                 row.names=NULL, stringsAsFactors = F)
installedSoftware
```


```{r,echo=FALSE,results='hide',message=FALSE}
 devtools::load_all(pkg=".")
```

Load the `rspeaksnonmem` package
```{r}
library(rspeaksnonmem)
```

Copy an example dataset and model to a directory of your choice

```{r,results='hide',echo=T}
getwd()

file.copy(from = file.path(system.file("exdata", package="rspeaksnonmem"),"warfarin_PK_CONC_MKS.ctl"),
            to = getwd(), overwrite=T )
file.copy(from = file.path(system.file("exdata", package="rspeaksnonmem"),"warfarin_conc_pca.csv"),
            to = getwd(), overwrite=T )
```

```estimate_NM``` runs NONMEM for a given modelFile. The ```installInfo``` argument holds configuration information
for each tool (shown above) which is used to create the command for the system prompt. The ```cleanup``` argument deletes
ancillary files from the working directory.

```{r NONMEM }
estimate_NM(modelFile="warfarin_PK_CONC_MKS.ctl", installInfo=installedSoftware, cleanup=T)
```

```{r Sumo }
sumo_PsN(listFile="warfarin_PK_CONC_MKS.lst", installInfo=installedSoftware)
```

Look at the output
```{r,message=F}
basicGOF_Xpose(modfile="warfarin_pk_conc_mks.ctl", listfile="warfarin_pk_conc_mks.lst")
```

The `defineExecutable` function used by ```estimate_NM``` and PsN calling functions will search for specified command 
executables at the system prompt:
```{r defineExecutable }
defineExecutable(command = "execute-3.5.4")
```
or it can search for suitable executables in a specified path:
```{r defineExecutable2}
defineExecutable(command = "nmfe72", installPath = "c:/nm72/", searchCommand = T)
```

The PsN functions work in an analogous way. Here we use ```execute_PsN``` to run estimation, but using PsN's additional 
functionality for restarting, tweaking initial estimates etc. Additional arguments used for PsN can be passed through the
```addargs``` argument to the function:

```{r executePsN}
execute_PsN(modelFile='warfarin_PK_CONC_MKS.ctl', command="execute-3.5.4", addargs="-tweak_inits -retries=3")
```
Next we can perform VPC using PsN:
```{r VPC_PsN}
VPC_PsN(modelFile='warfarin_PK_CONC_MKS.ctl', installInfo=installedSoftware, 
              nsamp=100, seed=54321)
```

and then use the Xpose function xpose_vpc to plot the result:
```{r plot_VPC}

```

```{r bootstrap_PsN}
bootstrap_PsN(modelFile='warfarin_PK_CONC_MKS.ctl', installInfo=installedSoftware, 
              nsamp=20, seed=54321)
```

# some example text for vignettes

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
