---
title: "Using rspeaksnonmem to run NONMEM & PsN"
author: "Mike K Smith (Mike.K.Smith@Pfizer.com)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setKnitrOptions, echo=FALSE,}
  library(knitr)
  opts_chunk$set(fig.width=6, fig.height=6)
```
`rspeaksnonmem` is designed to allow the user to craft workflows based on a given NONMEM model.

`rspeaksnonmem` allows the user to run NONMEM models from R either directly running the nmfe.bat or by
calling [Perl speaks NONMEM](http://psn.sourceforge.net) (PsN) functions like "execute". Other PsN functions can be run from
R - VPC (Visual Predictive Check), bootstrap, SSE (Stochastic Simulation and Estimation).

This allows the user to define sequences of pharmacometrics workflow tasks via R within a single R script. This increases traceability and reproducibility of workflow.

## Installation 

### Dependencies 

`rspeaksnonmem` relies on the package [`RNMImport`](https://github.com/MangoTheCat/RNMImport) function 
`importNmMod` which reads and parses the NONMEM control stream. `rspeaksnonmem` then works with
the data, parameter values, and task information separately from the model.

Before installation of `rspeaksnonmem`, the package `RNMImport` needs to be installed. This package is not on CRAN but is available on R-Forge. You can install from there using:

```{r install_RNMImport, include=F}
  #devtools::install_github("MikeKSmith/RNMImport")
```

### Install rspeaksnonmem
Eventually, `rspeaksnonmem` will be released to CRAN, but while still in development `rspeaksnonmem` can most easily be installed from GitHub using the `devtools` package:

```{r install_rspeaksnonmem, include=F}
    #devtools::install_github("MikeKSmith/rspeaksnonmem")
```

## Setup 
In order to execute NONMEM or PsN `rspeaksnonmem` needs to know where these tools are installed on your system.
The function defineExecutable creates the system command line for each tool by either referencing the tool name against 
an `installInfo` configuration file consisting of tool name, installation path and command e.g. "nonmem", "c:/nm72", "nmfe72.bat"; alternatively by searching for whether a command is configured on the system path; or by simply searching for executable files below an installation directory.

On Linux systems the "command" argument should provide the full path to the executable or define the command applicable
for running the tool at the system prompt. On Linux systems the "path" information on the configuration file may be left 
blank.

```{r setExecutables}
perlPath <- "c:/perl516"
PsNVersion <- "3.5.4"
NONMEMcommand <- "nmfe72"

tools <- c("NONMEM", "execute", "VPC", "bootstrap", "SSE", "runrecord", "sumo")
installPaths <- c("c:/nm72", rep(perlPath,length(tools)-1))
commands <- sapply(tools, function(x) ifelse( x == "NONMEM", NONMEMcommand, paste( casefold(x,upper = FALSE), PsNVersion, sep = "-" ) ) )

installedSoftware <- data.frame( cbind( tool = tools, path = installPaths, command = commands ), 
                                 row.names=NULL, stringsAsFactors = F)
installedSoftware
```


```{r,echo=FALSE,results='hide',message=FALSE}
 devtools::load_all(pkg=".")
 devtools::load_all(pkg = "C:\\Users\\smith_mk\\Documents\\Working documents\\RNMImport")
```

Load the `rspeaksnonmem` package
```{r, include=FALSE, echo=FALSE}
#library(rspeaksnonmem)
```

Copy an example dataset and model to a directory of your choice

```{r,results='hide',echo=T}
getwd()

file.copy(from = file.path(system.file("inst/exdata/", package="rspeaksnonmem"),"THEO.dat"),
            to = getwd(), overwrite=T )
file.copy(from = file.path(system.file("inst/exdata", package="rspeaksnonmem"),"Theophylline.mod"),
            to = getwd(), overwrite=T )
```

## Running NONMEM 
```estimate_NM``` runs NONMEM for a given model file. The ```installInfo``` argument holds configuration information
for each tool (shown above) which is used to create the command for the system prompt. The ```cleanup``` argument deletes
ancillary files from the working directory.

```{r NONMEM }
estimate_NM(modelFile = "theophylline.mod", installInfo = installedSoftware)
```
## Returning results from NONMEM output 
```{r Sumo}
sumo_PsN(lstFile = "theophylline.lst", installInfo = installedSoftware)
```

We can also use RNMImport to read the NONMEM output:

```{r RNMImport_import}
#library(RNMImport)
model1<-importNm(conFile="theophylline.mod", reportFile = "theophylline.lst")
```
 
and use the get<...> functions to extract information from the output into R objects. 
First let's retrieve the Objective Function Value (OFV): 
```{r getObjective}
getObjective(model1)	
```

And structural parameter estimates (THETAS): 
```{r getThetas}
getThetas(model1)
```

We can also return the initial, final and std errors of these parameters (stderror if the $COV step completes successfully): 
```{r getThetas2}
getThetas(model1, what = c("initial","final","stderrors")) 
```

We can extract OMEGA values as with getThetas: 
```{r getOmegas}
getOmegas(model1) 
```

We can extract the residual unexplained variability with getSigmas, analogously to getThetas: 
```{r getSigmas}
getSigmas(model1) 
```

Again, similarly, the covariance matrix of parameter estimates is available with getEstimateCov.
```{r getEstimateCov}
# The argument corMatrix can optionally return the correlation matrix.
getEstimateCov(model1, corMatrix = TRUE)
```

We can perform model diagnostics using the Xpose R package. This package requires certain output from the NONMEM control stream, namely the sdtab table including predictions and residuals; the patab file containing parameter estimates; and optionally the cotab and catab files containing continuous and categorical covariates respectively. Please see the Xpose documentation for more details. The basicGOF function gives a run summary and basic goodness of fit plots. It assumes that the run number is appended to the end of each of the table file names e.g. sdtab0, sdtab1 etc.

## Model qualification 
```{r,message=FALSE, warning=FALSE}
basicGOF_Xpose()
```

## Different ways of specifying the command to be run
The `defineExecutable` function used by ```estimate_NM``` and PsN calling functions can search for user-specified command 
to find valid executables at the system prompt:

```{r defineExecutable_with_command }
defineExecutable(command = "execute-3.5.4")
```

or it can search for suitable executables in a specified path:
```{r defineExecutable2}
defineExecutable(command = "nmfe72", installPath = "c:/nm72/", searchCommand = T)
```

The PsN functions work in an analogous way. Here we use ```execute_PsN``` to run estimation, but using PsN's additional 
functionality for restarting, tweaking initial estimates etc. Additional arguments used for PsN can be passed through the
```addargs``` argument to the function:

```{r executePsN}
execute_PsN(modelFile='theophylline.mod', command="execute-3.5.4", 
            psnOpts=list(tweak_inits=TRUE, retries=3))
```
## VPC 
Next we can perform VPC using PsN:
```{r VPC_PsN}
VPC_PsN( modelFile='theophylline.mod', installInfo=installedSoftware,
         working.dir="Run1_VPC", 
         samples=100, seed="54321",
         psnOpts=list(lst="theophylline.lst", bin_by_count=0,
                      bin_array="0.125,0.375,0.75,1.5,2.5,5,6,8,10,16,48"))
```

and then use the Xpose function xpose_vpc to plot the result:
```{r plot_VPC}
vpcdir <- "Run1_VPC"

vpctab <- file.path(vpcdir,"vpctab")
vpcinfo <- file.path(vpcdir,"vpc_results.csv")

xpose4::xpose.VPC(vpc.info=vpcinfo,	vpctab=vpctab)
```
## Bootstrap 
```{r bootstrap_PsN, include=FALSE}
# bootstrap_PsN(modelFile='theophylline.mod', installInfo=installedSoftware, 
#              nsamp=20, seed=54321, working.dir="Run1_bootstrap")
```
