---
title: "Using rspeaksnonmem"
author: "Mike Smith, Andrew C. Hooker and Ron Keizer "
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`rspeaksnonmem` is designed to allow the user to craft workflows based on a given NONMEM model.

After importing and parsing a control stream to an R object (using the `RNMImport` package), `rspeaksnonmem` allows the user to easily change initial estimates, change data attributes or change task properties (estimation or simulation settings) without having to change the model. This saves a lot of textual searching by the user. The resulting data, parameter,
model and task information is then written back out to file using an existing (possibly the original) model as a 
template.

`rspeaksnonmem` also allows the user to run NONMEM models from R either directly running the nmfe.bat or by
calling [Perl speaks NONMEM](http://psn.sourceforge.net) functions like "execute". Other PsN functions can be run from
R - VPC (Visual Predictive Check), bootstrap, SSE (Stochastic Simulation and Estimation).

## Installation

Eventually, `rspeaksnonmem` will be released to CRAN, but while still in development `rspeaksnonmem` can most easily be installed from GitHub using the `devtools` package:

    library(devtools)
    install_github("MikeKSmith/rspeaksnonmem")

## Dependencies

`rspeaksnonmem` relies on the package [`RNMImport`](https://r-forge.r-project.org/R/?group_id=1922) function 
`importNmMod` which reads and parses the NONMEM control stream. `rspeaksnonmem` then works with
the data, parameter values, and task information separately from the model.

Before installation of `rspeaksnonmem`, the package `RNMImport` needs to be installed. This package is not on CRAN but is available on R-Forge. You can install from there using:

    install.packages("RNMImport", repos="http://R-Forge.R-project.org")

or download the `.tar` file from [R-Forge](https://r-forge.r-project.org/R/?group_id=1922) and run in R:

    install.packages("~/Downloads/RNMImport_4.0-27.tar.gz", repos=NULL, type = "source")


## getting started
```{r setExecutables}
perlPath <- "c:/perl516"
PsNVersion <- "3.5.4"
installedSoftware <- list(NONMEM = list(path = "c:/nm72", command = "nmfe72"), 
                          execute = list(path = perlPath, command = paste("execute",PsNVersion,sep="-")), 
                          VPC = list(path = perlPath, command = paste("VPC",PsNVersion,sep="-")), 
                          bootstrap = list(path = perlPath, command = paste("bootstrap",PsNVersion,sep="-")), 
                          SSE = list(path = perlPath, command = paste("SSE",PsNVersion,sep="-")), 
                          runrecord = list(path = perlPath, command = paste("runrecord",PsNVersion,sep="-")))
```


```{r,echo=FALSE,results='hide',message=FALSE}
# devtools::load_all(pkg=".")
devtools::install(pkg=".")
```

Load the `rspeaksnonmem` package
```{r}
library(rspeaksnonmem)
```

Copy an example dataset and model to a directory of your choice

```{r,results='hide',echo=T}
user_dir <- "~/Working documents/rspeaksnonmem/Foo"
setwd(user_dir)
# dir.create(user_dir)
# file.copy(file.path(system.file("exdata", package="rspeaksnonmem")),
#           file.path(user_dir))
# myFilePath <- system.file("exdata", package="rspeaksnonmem")
```

Run the model:

```{r NONMEM}
estimate_NM(modelFile='warfarin_PK_CONC_MKS.ctl')
```

Look at the output
```{r,message=F}
cur_dir <- getwd()
setwd(user_dir)
basicGOF_Xpose()
setwd(cur_dir)
```

```{r executePsN}
execute_PsN(modelFile='warfarin_PK_CONC_MKS.ctl')
```
```{r bootstrapPsN}
bootstrap_PsN(modelFile='warfarin_PK_CONC_MKS.ctl')
```

# some example text for vignettes

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
