---
title: "Using rspeaksnonmem to modify NONMEM control streams"
author: "Mike Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`rspeaksnonmem` is designed to allow the user to craft workflows based on a given NONMEM model.

For more information on how to use `rspeaksnonmem` to run NONMEM and PsN please review the vignette 
"Using rspeaksnonmem to run NONMEM and PsN".

After importing and parsing a control stream to an R object using the `RNMImport` package, `rspeaksnonmem` allows the user to easily change initial estimates, change data attributes or change task properties (estimation or simulation settings) without having to change the model. This saves a lot of textual searching by the user. The resulting data, parameter,
model and task information is then written back out to file using an existing (possibly the original) model as a 
template. This ensures that the written NONMEM control stream mirrors as closely as possible the structure of the original file.

## Installation 

### Dependencies 

`rspeaksnonmem` relies on the package [`RNMImport`](https://github.com/MangoTheCat/RNMImport) function 
`importNmMod` which reads and parses the NONMEM control stream. `rspeaksnonmem` then works with
the data, parameter values, and task information separately from the model.

Before installation of `rspeaksnonmem`, the package `RNMImport` needs to be installed. This package is not on CRAN but is available on R-Forge. You can install from there using:

```{r install_RNMImport, include=F}
  #devtools::install_git("MangoTheCat/RNMImport")
```

### Install rspeaksnonmem
Eventually, `rspeaksnonmem` will be released to CRAN, but while still in development `rspeaksnonmem` can most easily be installed from GitHub using the `devtools` package:

```{r install_rspeaksnonmem, include=F}
    #devtools::install_github("MikeKSmith/rspeaksnonmem")
```


```{r,echo=FALSE,results='hide',message=FALSE}
 devtools::load_all(pkg=".")
 devtools::load_all(pkg="../../RNMImport")
```

Load the `rspeaksnonmem` package
```{r, include=FALSE, echo=FALSE}
#library(rspeaksnonmem)
```

Copy an example dataset and model to a directory of your choice

```{r,results='hide',echo=T}
getwd()

file.copy(from = file.path(system.file("exdata", package="rspeaksnonmem"),"warfarin_conc_pca.csv"),
            to = getwd(), overwrite=T )
file.copy(from = file.path(system.file("exdata", package="rspeaksnonmem"),"warfarin.ctl"),
            to = getwd(), overwrite=T )
```

The initial model within the workflow should act as a "template" for modifications. The best practice is to use a model where all possible parameters are defined (including OMEGAs and covariances / correlations) but where it is possible to fix parameters to zero or some null value. We can then run and test a wide variety of models simply by updating the $THETA, $OMEGA and $SIGMA parameters to allow estimation.

## Read the control stream using RNMImport 
First, we need to read the NONMEM control stream into R using the importNmMod function of RNMImport.
```{r readControlStream}
warfModel <- importNmMod("warfarin.ctl")
class(warfModel)
names(warfModel)
```

This creates a list of R objects containing the Raw NONMEM control stream as a vector of characters:
```{r viewRawNM}
warfBaseRaw <- warfModel$Raw
cat(warfModel$Raw, sep="\n")
```

It also creates a parsed list of control statements in the controlFile slot:
```{r viewParsedNM}
warfBaseModel <- warfModel$problemContents[[1]]
warfModel$problemContents[[1]]
```

Using `rspeaksnonmem` the user can then update elements of the model using this parsed set of commands. `rspeaksnonmem` provides some additional functions to extract certain elements of the parsed control stream. We view the $DATA and $INPUT commands as "Data" items; $THETA, $OMEGA and $SIGMA as "Parameter" items; $EST, $COV and $TAB as "Task" items; and everything else as "Model" items. 

NOTE: Generally speaking, `rspeaksnonmem` is set up to allow the user to quickly and easily update Data, Parameter or Task items, while leaving the Model unchanged. 

The `rspeaksnonmem` function `getNMDataObjects` returns the RAW $DATA and $INPUT lines from the NONMEM control stream and the associated parsed objects.

```{r getNMData}
getNMDataObjects(warfModel)
```
 
Similarly, `getNMParameterObjects` returns the $THETA, $OMEGA and $SIGMA records.
```{r getNMParameterObjects}
getNMParameterObjects(warfModel)
```
 
Finally, `getNMTasProperties` returns the $EST, $COV and $TABLES records.
```{r getNMTaskPropertiesObjects}
getNMTaskPropertiesObjects(warfModel)
```

Together the Data, Parameters, Model and Task information forms a "Modelling Object Group" or MOG which is used for a specific estimation task.
We can update the MOG using the `rspeaksnonmem` function `updateModel`.
```{r updateMOG}
newTheta <- warfBaseModel$Theta
newTheta["POP_TLAG",] <- list(Lower=0, Est=0.75, Upper=1.5, FIX=FALSE, comments="POP_TLAG")
newWarfModel <- updateModel(parsedObject = warfBaseModel, theta=newTheta)
newWarfModel
```

Note how the THETA settings for POP_TLAG have been updated, but all other elements are unchanged. It is also possible to change the initial estimates while retaining the specified lower and upper bounds:

```{r}
newThetas <- warfBaseModel$Theta
newInits <- c(0.25, 10, 4, 0.25, 0, 0.1)
newThetas[,"Est"] <- newInits

newModel <- updateModel(parsedObject = warfBaseModel, theta=newThetas)
newModel
```

We can finally write out the updated control stream using the function `writeNMControlStream`. This function uses an existing NONMEM control stream (or raw records) as a template for writing the updated, parsed statements. This ensures that the order of statements corresponds between the original and updated files, which allows easier checking of changes.

```{r writeNMControlStream}
writeNMControlStream(templateModel = warfModel$Raw, parsedControl = newModel, modelFile = "warfarin2", modelExtension = "ctl")
cat(readLines("warfarin2.ctl"), sep="\n")
```

