---
title: "Using rspeaksnonmem to run NONMEM & PsN"
author: "Mike K Smith (Mike.K.Smith@Pfizer.com)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
#output: html_notebook
---

```{r setKnitrOptions, echo=FALSE,}
  library(knitr)
  opts_chunk$set(fig.width=6, fig.height=6)
```
`rspeaksnonmem` is designed to allow the user to craft workflows based on a 
given NONMEM model.

`rspeaksnonmem` allows the user to run NONMEM models from R either directly 
running the nmfe.bat or by
calling [Perl speaks NONMEM](http://psn.sourceforge.net) (PsN) functions like 
"execute". Other PsN functions can be run from R - VPC (Visual Predictive Check),
bootstrap, SSE (Stochastic Simulation and Estimation).

This allows the user to define sequences of pharmacometrics workflow tasks via 
R within a single R script. This increases traceability and reproducibility of 
workflow.

```{r setup,echo = FALSE, results = 'hide', message = FALSE, warning = FALSE}
devtools::load_all(pkg=".")
devtools::load_all(pkg = "C:\\Users\\smith_mk\\Documents\\Working documents\\RNMImport")

getwd()

file.copy(from = file.path(system.file("inst/exdata/", package="rspeaksnonmem"),
                           "warfarin_conc_pca.csv"),
          to = getwd(), overwrite = T )
file.copy(from = file.path(system.file("inst/exdata", package="rspeaksnonmem"),
                           "warfarin.ctl"),
          to = getwd(), overwrite = T )

```

### Running NONMEM 
`estimate_NM` runs NONMEM for a given model file. Note 
that in the `command` argument to the function we specify the complete path to the
NONMEM executable .bat file. This allows you to be very precise about which
version of NONMEM you use. If the `clean` argument is non-zero it cleans up the 
working directory deleting all of the temporary NONMEM files, leaving only the 
output. 

```{r NONMEM , message = FALSE, warning = FALSE}
estimate_NM(modelFile = "warfarin.ctl", 
            command = "c:/nm72/run/nmfe72.bat", 
            clean = 1)
```

### Running PsN
We can now use PsN command `sumo` to summarise the output. Note that here we
are not using the `command` argument, but instead we specify the installation
directory for Perl / PsN and which PsN version we are using. `rspeaksnonmem` 
will then create a command based on the directory, `sumo` and version information.

```{r callSumo}
  sumo_PsN(installPath = "c:/strawberry/perl",
           version = "4.6.0",
           lstFile = "warfarin.lst")
```

### Different ways of specifying the command to be run
The `defineExecutable` function used by `estimate_NM` and PsN calling 
functions can search for user-specified command to find valid executables at 
the system prompt. If you supply the command with no file path with 
searchCommand=T then the function uses `Sys.which` or `Sys.which2` (depending 
on OS) to search the system PATH or shortcut definitions for executables.

```{r defineExecutable_with_command }
defineExecutable(command = "R", searchCommand = T)
```

If you specify the `tool` and the `installPath`, then `defineExecutables` looks 
within the `installPath` directory for executables that match the tool name.

```{r defineExecutable2}
defineExecutable(tool = "nonmem", installPath = "c:/nm72", searchCommand = T)
```

Note that for PsN there could are a few ways of invoking the routine. 
```{r defineExecutable3}
defineExecutable(tool = "vpc", 
                 installPath = "c:/strawberry/perl", 
                 searchCommand = T)
```

In the `rspeaksnonmem` calls to PsN routines we explicitly tie the Perl script 
to the "parent" Perl executable. We use the version number Perl script so that 
it's clear to anyone looking at the command which version of the Perl script 
has been used and the Perl executable associated with it.

```{r defineExecutable4}
defineExecutable(tool = "vpc", 
                 installPath = "c:/strawberry/perl", 
                 version = "4.6.0")
```

### Validating PsN option lists
`rspeaksnonmem` has functionality to check that the name-value pairs specified
in the psnOpts arguments of the PsN calling functions are appropriately named
and have valid argument types for use with PsN. We do this by checking against
the routines' `-h` help files.

```{r}
  psnOpts <- list(picky = "yes",  ## retries option should be numeric
                  auto_bin = 4,     ## auto_bin is not a valid option for execute
                  dir = getwd())

  validate_PsN_options(tool = "execute",
                       installPath = "c:/strawberry/perl",
                       version = "4.6.0",
                       psnOpts = psnOpts)
```

### Calling PsN
`rspeaksnonmem` has PsN call functions for the following routines:
  
  - `execute`
  - `sumo`
  - `vpc`
  - `bootstrap`
  - `sse`
  - `runrecord`
  
### Other PsN functionality
Although functions have been provided as wrappers to the PsN functions, it is 
easy to extend these to any other PsN function using the function callPsN.
```{r other_PsN}
  callPsN(tool = "linearize", 
          installPath = "c:/strawberry/perl",
          version = "4.6.0",
          file = "warfarin.ctl")
```
